class: inverse, center, middle
# RStudio for Basic Statistics


---
# Topics covered

- Fisher's exact test and chi-squared test for contingency table data
- Wilcoxon rank sum test and Kruskal-Wallis test for non-normally distributed continuous data
- T-test and ANOVA for normally distributed continuous data
- Wilcoxon signed rank test and paired t-test for paired data


---
# Load the contingency table data

```{r}
cat1 <- readxl::read_excel(
  here::here("rstudio-demo", "data", "StatsDataSet-1A.xlsx"))
cat1
```

Notes:

- R requires column names for all data columns, so you'll see a message stating that "...1" was automatically supplied for the column that didn't have a name. 
- By default R considers anything in the top row to be column names and anything in following rows to be data.


---
# Load the contingency table data

First, let's rename that column so that it has a useful name:

```{r message = FALSE}
library(dplyr)
cat1 <- rename(cat1, Country = ...1)

cat1
```


---
# Convert to a matrix for testing

In this case, with aggregated data, we need to convert this 2x2 table to a matrix of numeric values to apply the tests of interest.

This basically just means removing the first column that contains the country, and converting from a dataframe (or tibble) to a matrix.

```{r}
cat1_mat <- as.matrix(cat1[, 2:3])

cat1_mat
```


---
# Tips

- The rows and columns of a dataframe can be referenced by name or position within square brackets `[rows, columns]`
- So `cat1[1, 2]` refers to the first row and second column
- On the previous slide, `cat1[, 2:3]` refers to all rows (first argument left blank) and columns 2 through 3
- Non-consecutive rows or columns could be referenced as, for example, `cat1[, c(1, 3)]` for all rows and the first and third column, where `c()` denotes a vector
- `cat1[1, "Red hair"]` would give the "Red hair" column for the first row (i.e. Country = Irish)


---
# Fisher's exact test

We can test the null hypothesis of no association between country and hair color by performing a Fisher's exact test.

```{r}
fisher.test(x = cat1_mat)
```


---
# Chi-squared test

Alternatively, we can use a chi-squared test, since all expected cell counts are >=5.

```{r}
chisq.test(x = cat1_mat)
```


---
# Tips

- Arguments are processed in order in R functions. 
- You can supply the arguments in any order if you specify the argument names.
- If you do not specify the argument names, the arguments must be supplied in order

Here are all of the arguments available for the `fisher.test()` function:

```{r eval = FALSE}
fisher.test(x, y = NULL, workspace = 200000, hybrid = FALSE,
            hybridPars = c(expect = 5, percent = 80, Emin = 1),
            control = list(), or = 1, alternative = "two.sided",
            conf.int = TRUE, conf.level = 0.95,
            simulate.p.value = FALSE, B = 2000)
```

So `fisher.test(x = cat1_mat)` and `fisher.test(cat1_mat)` are equivalent.

If you wanted to change to a one-sided alternative:

- `fisher.test(cat1_mat, alternative = "greater")` works
- `fisher.test(x = cat1_mat, alternative = "greater")` works
- `fisher.test(alternative = "greater", x = cat1_mat)` works
- `fisher.test(cat1_mat, "greater")` will not work, since `alternative` is not the second argument.


---
# Load the continuous data with two groups

These data are formatted with two columns - Sex and HDL - and each row contains the data for one subject.

```{r}
cont1 <- readxl::read_excel(
  here::here("rstudio-demo", "data", "StatsDataSet-2B.xlsx")
  )

cont1
```


---
# Visually check the distribution

For such a small sample size (n=24) we would not expect normality to hold - we usually require at least 30 samples to perform tests that rely on the normality assumption. However, we can examine a histogram to check.

```{r fig.width = 5, fig.height = 5}
hist(cont1$HDL)
```


---
# Wilcoxon rank-sum test

a.k.a. Mann-Whitney test

We can test the null hypothesis of no association between sex and HDL.

```{r}
wilcox.test(HDL ~ Sex, data = cont1)
```

The warning message tells us that there are ties, or observations with the same value, in the data. So a normal approximation will be used instead of an exact p-value.


---
# T-test

For demonstration purposes, we will also apply the t-test to these data.

```{r}
t.test(HDL ~ Sex, data = cont1)
```


---
# Tips

- For the preceding two tests we used the `formula` option to supply the information for the test
- A formula in R looks like `LHS ~ RHS` where the left-hand side is typically an outcome variable or dependent variable and the right-hand side is the independent variable.
- For multivariable regression, you can specify formulas in the format `Y ~ X1 + X2 + X3` when there is more than one independent variable.
- You must also supply the name of the dataset to the `data = ` argument so that R knows where to look to find the specified variables.
- R is case sensitive so "sex" is not the same as "SEX" or "Sex". This holds true for both variable names and variable values - e.g. "Female" and "female" would be considered different values for sex


---
# Load the continuous data for >2 groups

These data are formatted with two columns - Strain and Atherosclerosis - and each row contains the data for one subject.

```{r}
cont2 <- readxl::read_excel(
  here::here("rstudio-demo", "data", "StatsDataSet-3.xlsx"))
cont2
```


---
# Kruskal-Wallis test

A generalization of the Wilcoxon rank-sum/Mann-Whitney test for >2 groups

```{r}
kruskal.test(Atherosclerosis ~ Strain, cont2)
```


---
# ANOVA

For demonstration purposes we will also perform ANOVA (without checking the normality assumption).

```{r}
lm_mod <- lm(Atherosclerosis ~ Strain, cont2)
anova(lm_mod)
```


---
# Load the paired continuous data

HDL levels for each subject before and after treatment with some drug.

```{r}
cont3 <- readxl::read_excel(
  here::here("rstudio-demo", "data", "StatsDataSet-2C.xlsx"))
cont3
```


---
# Wilcoxon signed-rank test

We test the null hypothesis that there is no difference in HDL before versus after treatment.

```{r}
wilcox.test(cont3$`Before drug`, cont3$`After drug`, paired = TRUE)
```


---
# Paired t-test

For demonstration purposes we also try the paired t-test without checking the normality assumption.

```{r}
t.test(cont3$`Before drug`, cont3$`After drug`, paired = TRUE)
```


---
# Tips

- We always need to tell R where to find a variable.
- If the function doesn't have a `data` argument available, then we need to use the format `dataset$variable`, which tells R to look in the dataset specified before the $ to find the variable.
- Variable names with spaces must be referenced in back ticks or double quotes.
- I prefer to "clean up" my variable names to avoid this using the `clean_names()` function from the {janitor} package.

```{r message = FALSE}
library(janitor)
cont3_clean <- clean_names(cont3)
names(cont3)
names(cont3_clean)
```

















